---
title: "Equity Premium - Exploratory data analysis"
author: "Druce Vertes"
date: "July 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initialize

Libraries to use
```{r, echo=TRUE}
library(ggplot2)
library(lattice)
library(caret)
library(corrplot)
library(car)
library(RColorBrewer)


```
## Import and clean data

* Data from Amit Goyal's equity premium dataset

* http://www.hec.unil.ch/agoyal/

* http://www.hec.unil.ch/agoyal/docs/PredictorData2016.xlsx

* Import data from CSV

```{r}
setwd("C:/Users/druce/R/EquityPremium2017")
data<-read.csv('PredictorData2016q.csv',na.strings = c("NA","#DIV/0!", "","NaN"))

```

## Clean...Trim NA valued columns

```{r}
countMissing <- function(mycol) {
  return (sum(is.na(data[, mycol]))/ nrow(data))
}
countNAs <- data.frame(countNA=sapply(colnames(data), countMissing))
countNAs
subset(countNAs, countNAs$countNA > 0.5)
colsToDeleteNA <- countNAs$countNA > 0.5
data <- data[, !colsToDeleteNA]
```

## Clean...Trim NA valued rows

```{r}
rowsToDelete <- data$yyyyq <= 19254
data <- data[!rowsToDelete,]
```

## Add EqPrem column

```{r}
data$EqPrem = data$CRSP_SPvw - data$Rfree
```

## Explore data

```{r}
nearzero <- nearZeroVar(data, saveMetrics=TRUE)
nearzero
```

## Correlations of raw data

```{r}
corrdata <- data[,c("D12","E12","b.m","tbl","AAA","BAA","lty","ntis","Rfree","infl","ltr","corpr","svar","EqPrem")]
corrdata <- corrdata[complete.cases(corrdata),]
corrdata$D12log <- log(corrdata$D12)
corrdata$E12log <- log(corrdata$E12)
corrdata <- corrdata[,c("D12log","E12log","b.m","tbl","AAA","BAA","lty","ntis","Rfree","infl","ltr","corpr","svar","EqPrem")]
mycor <- cor(corrdata)
corrplot.mixed(mycor, upper="color", order="hclust")
```

## Correlations of first differences (and raw EqPrem)

```{r}
data2 <- data[,c("D12","E12","b.m","tbl","AAA","BAA","lty","ntis","Rfree","infl","ltr","corpr","svar","EqPrem")]
data2$D12log <- log(data2$D12)
data2$E12log <- log(data2$E12)
data2 <- data2[,c("D12log","E12log","b.m","tbl","AAA","BAA","lty","ntis","Rfree","infl","ltr","corpr","svar","EqPrem")]
diffs <- tail(data2, -1) - head(data2, -1)
diffs$EqPrem <- tail(data$EqPrem, -1)
corrdata <- diffs[,c("D12log","E12log","b.m","tbl","AAA","BAA","lty","ntis","Rfree","infl","ltr","corpr","svar","EqPrem")]
corrdata <- corrdata[complete.cases(corrdata),]
mycor <- cor(corrdata)
corrplot.mixed(mycor, upper="color", order="hclust")
```

## We see Rfree is perfectly correlated with tbl so we can remove it from further correlation analysis

```{r}
corrdata <- corrdata[, !(names(corrdata) %in% c("Rfree"))]
1
```
## Look at interesting correlations of a few variables with leads and lags

* dividends and earnings show significant positive serial correlation (if it's ttm then consecutive quarters share 9 months; but also we expect management to smooth and keep on trend.)

* might have thought EqPrem would have some modest serial correlation, but no

* some of the interest rates show odd patterns, almost wonder if there's an artifact like month-end vs. average


```{r}
attach(corrdata)
ccf(D12log, D12log, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(D12log, E12log, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(E12log, E12log, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(b.m, b.m, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(b.m, BAA, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(tbl, tbl, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(tbl, AAA, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(tbl, BAA, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(tbl, lty, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(tbl, ltr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(tbl, corpr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(AAA, AAA, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(AAA,  BAA, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(AAA, lty, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(AAA, ltr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(AAA, corpr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(BAA, BAA, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(BAA, lty, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(BAA, ltr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(BAA, corpr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(lty, lty, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(lty, ltr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(lty, corpr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(ntis, ntis, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(infl, infl, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(ltr, ltr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(ltr, corpr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(corpr, corpr, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(svar, svar, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")
ccf(EqPrem, EqPrem, ylab="correlation", lag.max=4, ylim=c(-1,1), col="red")

```
### All pairwise correlations of first differences (and raw EqPrem) with leads and lags

* above are the ones that look interesting, all below for reference)

```{r}
# cartesian product of column numbers
mygrid = expand.grid(c(1:ncol(corrdata)), c(1:ncol(corrdata)))
#make upper triangular
#commented out because I don't know how to make ccf skip empty slots
#mygrid <- mygrid[!(mygrid$Var1<mygrid$Var2),]

# function that takes 2 column numbers, runs ccf on corresponding 2 cols in corrdata
myccf <- function(i, j) {
      iname=colnames(corrdata)[i]
      jname=colnames(corrdata)[j]
      mytitle = paste(iname, jname, sep=" vs. ")
#      if (i<j) {
#        plot.new()        
#      }
#      else {
        ccf(corrdata[,i], corrdata[,j], ylim=c(-1,1), ylab="",xlab="", main=iname, lag.max=4, col="red") 
        if (i==1 ) {
          text(labels=jname, x=-1.75, y=-0.5)
        }
        if (j==1 ) {
          text(labels=iname, x=0, y=0.75)
          }      
      #}
}
# run charts of correlations with leads and lags
# run in grid, # no axis labels, # no margins
```
```{r, echo=FALSE}

old.par<-par(mfrow = c(ncol(corrdata), ncol(corrdata)), yaxt='n', xaxt='n', mar=c(0.1,0.1,0.1,0.1)) 
mapply(myccf, mygrid$Var1, mygrid$Var2)

```
## All charts from grid full size for reference
```{r, echo=TRUE}

#grid has limited usefulness without improving aesthetics...run all sequentially
myccf <- function(i, j) {
      iname=colnames(corrdata)[i]
      jname=colnames(corrdata)[j]
      mytitle = paste(iname, jname, sep=" vs. ")
      ccf(corrdata[,i], corrdata[,j], ylab="", xlab="", main=mytitle, lag.max=4)
}

#make upper triangular
mygrid <- mygrid[!(mygrid$Var1<mygrid$Var2),]
par(old.par)
mapply(myccf, mygrid$Var1, mygrid$Var2)

```
